Extreme climatic events with asymmetric tail-dependence affect population-extinction risk

### 2019-07-15 MainText.Rmd
writing intro

<!--how common are the extreme events?-->
For past few decades, there have been increasing evidences that anthropogenic climate change 
is causing extreme events, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness. Such extreme climatic events (ECE) can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; 
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) through 
life-history traits and this alteration in community structures could cascade 
through higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. For example, 
a study based on annal survival data showed changes
in prey communities induced by extreme flooding affected predators in 
aquatic and adjacent riparian habitats [@chiu2013impact]. Similarly, 
$43$ years of ring-recovery data was used to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland [@frederiksen2008demographic].
A recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years (1976-2012) at the site level, across 
each species' life-stages. Successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different stage class of a population making it vulnerable 
to local extinction.

<!--How copula could be relevent in estimating extreme event's influence on extinction risk : introduce Taildep. concept-->

<!--Tail dependence of copula and recent result from BIVAN showing LTdep with higher extrisk than UTdep, but 
not explored in age structured model or in metapopulation context with density dependence effect-->

<!--our objective ?-->
We explored the potential use of asymmetric tail-dependence in copula 
for understanding the extinction risk of populations towards extreme climatic event in ecology
by addressing the following questions: (Q1) Does nature of tail dependence in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does asymmetric tail-dependent noises across multiple patches influence extinction 
risk of a nonlinear metapopulation model with density dependence? Our speculation is that the general finding 
from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were 
left-tail dependent than the one if noises had right-tail dependence) holds true only in
the under-compensatory regime, but should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

We found density-dependence metapopulation models (with $n=5$ patches) in noisy 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in earlier studies [BIVAN paper]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher (or lower) extinction risk for 
the metapopulations when noises across multiple patches were left- (or right) tail dependent.
But that observation remained valid when the model parameter, usually growth rate $(r)$, of each model was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured. When the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$, we observed right (or left) tail dependent noise caused higher (or lower) extinction 
risk for the metapopulations. Figs. \ref{fig_extrisk_allmodel_localdisp} shows variation of extinction risk 
against local dispersal rate ($d$) for six density dependent metapopulation models 
where growth rate $r$ was taken either from under-compensatory or from over-compensatory 
regime. We also got similarly consisent results with 
our prediction (see Q2 from \nameref{intro}) for global 
dispersal (Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6}) 
shows how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) 
represents the same but when global dispersal took place between patches. 
Both case showed the signature of 
"rescue-effect" (i.e. by lowering extinction risk) when there was 
a finite dispersal compared to the one with 
$d=0$.

>>>>

<!--how common are the extreme events?-->
For past few decades, there have been increasing evidences that anthropogenic climate change 
is causing extreme events, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness. Such extreme climatic events (ECE) can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) through 
life-history traits and this alteration in community structures could cascade 
through higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. For example, 
a study based on annal survival data showed changes
in prey communities induced by extreme flooding affected predators in 
aquatic and adjacent riparian habitats [@chiu2013impact]. Similarly, 
$43$ years of ring-recovery data was used to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland [@frederiksen2008demographic].
A recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years (1976-2012) at the site level, across 
each species' life-stages. Successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction.

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] showed the importance, causes 
and consequences of copula structures in associations between ecologically
important variables and demonstrated one example of density-independent metapopulation model where 
copula was used to estimate extinction risk. Copula, a rank based measure, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. We will hereby illaborate the idea why
the potential use of copula in modelling extinction risk goes beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
<!--do I need to explain the details how to make a copula from raw data?-->

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is associations between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. smaller values 
(respectively, the right or upper portions, i.e. higher values) 
of the two distributions. For example, extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal), which 
could lead to population crash and thereby inflate the extinction risk. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex depending 
on the dynamics of the model. Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our knowledge
copula have never been used in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-dependence in a copula 
for understanding the extinction risk of populations towards extreme climatic event in ecology
by addressing the following questions: (Q1) Does nature of tail dependence 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does asymmetric tail-dependent noises across multiple patches influence extinction 
risk of a nonlinear metapopulation model with density dependence? Our speculation is that the general finding 
from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were 
left-tail dependent than the one if noises had right-tail dependence) holds true only in
the under-compensatory regime, but should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

<!--Tuljapurkar estimation to be explained in Discussion-->

@@

We found density-dependence metapopulation models (with $n=5$ patches) in noisy 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in earlier studies [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher (or lower) extinction risk for 
the metapopulations when noises across multiple patches were left- (or right) tail dependent.
But that observation remained valid when the model parameter, usually growth rate $(r)$, of each model was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured. When the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$, we observed right (or left) tail dependent noise caused higher (or lower) extinction 
risk for the metapopulations. Figs. \ref{fig_extrisk_allmodel_localdisp} shows variation of extinction risk 
against local dispersal rate ($d$) for six density dependent metapopulation models 
where growth rate $r$ was taken either from under-compensatory or from over-compensatory 
regime. We also got similarly consisent results with 
our prediction (see Q2 from \nameref{intro}) for global 
dispersal (Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6}) 
shows how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) 
represents the same but when global dispersal took place between patches. 
Both case showed the signature of 
"rescue-effect" (i.e. by lowering extinction risk) when there was 
a finite dispersal compared to the one with 
$d=0$.
